Hybrid Threads \\ for the \\ Parrot Virtual Machine

### 2012-06-26 Paper.tex
First diagrams appear

\usepackage{wrapfig}
\usepackage{listings}
\AddBibFile{literature.bib}

@@

\section{PolyMorphic Containers (PMCs)}
\label{sec:PMCs}

@@

A runloop is the inner most loop executing bytecode in Parrot. It is also a data structure containing some information needed to support exception handling. When some operation in Parrot bytecode is calling a C function and this C function in turn is again executing Parrot bytecode, a nested runloop is started. Examples for such situations include calling a library function with a callback as parameter and exceptions thrown inside Parrot's C code calling a previously defined exception handler in user code.

@@

As described in the introduction, cross-thread writes to shared variables may endanger the internal consistency of the interpreter. Traditionally, the solution to this problem is the use of locks of varying granularity. Fine-grained locking allows code to run in parallel but taking and releasing locks costs performance. It not only increases the instruction count and memory accesses but it also forces the CPU cores to coordinate and thus communicate. Even a seemingly simple operation like an atomic increment can take two orders of magnitude longer than a normal increment \cite{LockingInOSKernels}. While the gain through being able to utilize multiple CPU cores may offset this cost, it is still impacting the common case of having only a singlethread running.

@@

The solution for these problems implemented in this thesis is to side step them altogether by disallowing write access to shared variables. The programmer (or in most cases the compiler) is obliged to declare a list of all shared variables before a newly created task is started. The interpreter would then create proxy objects for these variables which the task can use to access the data. These proxies contain references to the original objects. They use these references to forward all reading \textit{vtable} functions to the originals. Write access on the other hand would lead to a runtime error.

@@

>>>>

\usepackage{wrapfig}
\usepackage{listings}
\usepackage{tikz}
\usepackage{pgf-umlsd}
\usepgflibrary{arrows}
\usetikzlibrary{shadows}
\AddBibFile{literature.bib}

@@

\section{\textit{PolyMorphic Containers} (PMCs)}
\label{sec:PMCs}

@@

A runloop is the inner most loop executing bytecode in Parrot. It consists of two steps as shown in the following pseudo code:
\begin{GenericCode}
while (pc != NULL) { // program counter points to an opcode
    op = fetch_opcode_at(pc);
    pc = execute_opcode(op);
}
\end{GenericCode}

\begin{figure}
\centering
\includegraphics[scale=.4]{runloops}
\caption{Example of a callback executing in a nested runloop.}
\label{fig:runloops}
\end{figure}

It is also a data structure containing information needed to support exception handling. When some operation in Parrot bytecode is calling a C function and this C function in turn is again executing Parrot bytecode, a nested runloop is started. Examples for such situations include calling a library function with a callback as parameter and exceptions thrown inside Parrot's C code calling a previously defined exception handler in user code. This is shown in figure~\ref{fig:runloops}.

@@

As described in the introduction, cross-thread writes to shared variables may endanger the internal consistency of the interpreter. Traditionally, the solution to this problem is the use of locks of varying granularity. Fine-grained locking allows code to run in parallel but taking and releasing locks costs performance. It not only increases the instruction count and memory accesses but it also forces the CPU cores to coordinate and thus communicate. Even a seemingly simple operation like an atomic increment can take two orders of magnitude longer than a normal increment \cite{LockingInOSKernels}. While the gain through being able to utilize multiple CPU cores may offset this cost, it is still impacting the common case of having only a single thread running.

@@

The solution for these problems implemented in this thesis is to sidestep them altogether by disallowing write access to shared variables. The programmer (or in most cases the compiler) is obliged to declare a list of all shared variables before a newly created task is started. The interpreter would then create proxy objects for these variables which the task can use to access the data. These proxies contain references to the original objects. They use these references to forward all reading \textit{vtable} functions to the originals. Write access on the other hand would lead to a runtime error.

@@

\begin{figure}
    \centering
    \begin{sequencediagram}
        \newthread{main}{:MainThread}
        \newinst{shared}{:A}
        \newthread{child}{:ChildThread}

        \begin{callself}{main}{task1}{}
        \end{callself}

        \prelevel\prelevel
        \begin{callself}{child}{task2}{}
            \postlevel
            \begin{messcall}[1]{child}{schedule(write\_task)}{main}
                \begin{callself}{main}{write\_task}{}
                    \begin{messcall}{main}{write}{shared}
                    \end{messcall}
                \end{callself}
            \end{messcall}
        \end{callself}

        \prelevel\prelevel
        \begin{callself}{main}{task1}{}
        \end{callself}

    \end{sequencediagram}
    \caption{\textit{task2} running on \textit{ChildThread} is sending the \textit{write\_task} to the \textit{MainThread} to write to the shared variable \textit{A}.}
    \label{fig:hybridthreads}
\end{figure}
